---
title: "Expedia case study"
author: "Joan Biel"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---
<!--
rmarkdown::run("analysis_expedia_case_study.Rmd", shiny_args = list(port = 8080,launch.browser =FALSE))
app.files <- list('analysis_expedia_case_study.Rmd', 'www/expedia_casestudy_20170127.csv',"www/aggregates.rda","www/bookings_fit.rda")
shinyapps::deployApp(appFiles=app.files,appName='expedia_case_study',account="jibietr",launch.browser =FALSE)
-->


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(plyr)
library(reshape)
```


### Basic statistics

```{r echo=FALSE, message=FALSE, warning=FALSE}
PATH <- "www/"
fname <- paste(PATH,'expedia_casestudy_20170127.csv',sep='')
bookings <- read.csv(fname,sep=",",header=TRUE,fill=TRUE)
colnames(bookings) <- c("partner_id","mkt","bkg_value")
bookings$bkg_id <- seq(nrow(bookings))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
load(file="www/aggregates.rda")
```



#### Total number of bookings in data

```{r echo=FALSE, message=FALSE, warning=FALSE}
nrow(bookings)
```

#### Num partners
```{r echo=FALSE, message=FALSE, warning=FALSE}
partners <- levels(aggr.prt.mkt$partner_id)
length(partners)
```
#### Num markets
```{r echo=FALSE, message=FALSE, warning=FALSE}
markets <- levels(aggr.prt.mkt$mkt)
length(markets)
```

#### Booking value

Basic stats of booking value:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
summary(bookings$bkg_value)
```

As shown in the figure below, the distribution of bookings value is bimodal in the log scale.
This seems to indicate two different type of bookings or booking value groups.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(bookings,aes(x=log10(bkg_value)))+geom_histogram() + ylab("# bookings") + xlab("booking value (in log10)")
```

We can fit a univariate mixture of 2 gaussian components with variable variance
to identify the two modals in the booking value:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
library(mclust)
# fix a univariate mixture of 2 components and variable variance
#mod = densityMclust(log10(bookings$bkg_value+1e-15),2)
load(file="www/bookings_fit.rda")
plot(mod, what = "density", data = log10(bookings$bkg_value), breaks = 15,xlab="bkg_value (log10)",xlim=c(-6,1))
# I can use this to classify
bookings$class <- mod$classification
```

As a result, modal 1 has mean =`r mod$parameters$mean[1]` and SD = `r mod$parameters$variance$sigmasq[1]`,
whereas modal 2 has a mean `r mod$parameters$mean[2]` and SD = `r mod$parameters$variance$sigmasq[2]` (in the log scale).
In the linear scale, this correspond to means of `r 10^mod$parameters$mean[1]` and `r 10^mod$parameters$mean[2]` respectively,
i.e. a difference of two orders of magnitude.

**Note:** the mean statistic of 0.259 for the booking value (show above) comes roughly from averaging the two peaks
(this happens when the area under the curve of each model is similar).

### PART 1. Market analysis.


#### Volume vs Value

* Figure below shows the number of bookings vs the market value for each partner and market pair.

* There is a linear relationship between the demand/size of market (in number of bookings) and
the total market value generated (for some markets, these two metrics of market performance are non-linear).
In the next sections, we will use total value as market performance measure.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.prt.mkt,aes(x=num_bkgs,y=tot_prt_mkt_bkg_value))+geom_point() + xlab("Num bookings in market") + ylab("Partner's market value")  
```

#### Distribution of total booking value per market

* Distribution of total booking per value is right-tailed (see left plot below). Some markets concentrate a large number of total booking values. At the right, same plot in log axis.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.mkt,aes(x=tot_mkt_bkg_value))+geom_histogram() + ylab("# mkt") + xlab("total booking value")  
ggplot(aggr.mkt,aes(x=log10(tot_mkt_bkg_value)))+geom_histogram() + ylab("# mkt") + xlab("total booking value (log10)")  
```

* Basic statistics of the total value per market:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
summary(aggr.mkt$tot_mkt_bkg_value)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
aggr.mkt <- arrange(aggr.mkt ,desc(tot_mkt_bkg_value))
aggr.mkt$per_tot_mkt_bkg_value <- cumsum(aggr.mkt$tot_mkt_bkg_value)/sum(aggr.mkt$tot_mkt_bkg_value)
aggr.mkt$id <- seq(nrow(aggr.mkt))
num.top.markets <- sum(aggr.mkt$per_tot_mkt_bkg_value<.50)
```

As shown in the cumulative distribution below, `r num.top.markets` markets concentrate 25\% of the value generated to Expedia:

> **Result:** `r num.top.markets` markets concentrate 50\% of the value generated to Expedia:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.mkt, aes(x=id,y=per_tot_mkt_bkg_value)) + geom_point() + xlab("mkts") + ylab("% total booking value")
```

#### Distribution of partners per market

Figure below shows the distribution of partners per market.
Distribution is right-tailed. In average, there are `r mean(aggr.mkt$num_partners)` partners competing
in a market.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.mkt,aes(x=num_partners))+geom_histogram() + ylab("# mkt") + xlab("# partners")
summary(aggr.mkt$num_partners)
```

### PART 2: Partner analysis.

#### Markets per partner

Figure below shows the distribution of markets per partner. Distribution is right-tailed. In average, a partner is present in `r mean(aggr.prt$num_markets)` markets.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.prt,aes(x=num_markets))+geom_histogram() + ylab("# partners") + xlab("# markets")  
summary(aggr.prt$num_markets)
```

#### Total booking value per partner

* Figure below shows thd distribution of total booking per value is right-tailed (some partners accumulate large booking value).

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.prt,aes(x=tot_prt_bkg_value))+geom_histogram() + ylab("# partners") + xlab("total booking value")  
ggplot(aggr.prt,aes(x=log10(tot_prt_bkg_value)))+geom_histogram() + ylab("# partners") + xlab("total booking value (log10)")  
```

Basic stats of the total value per partner:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
summary(aggr.prt$tot_prt_bkg_value)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
aggr.prt <- arrange(aggr.prt ,desc(tot_prt_bkg_value))
aggr.prt$per_tot_prt_bkg_value <- cumsum(aggr.prt$tot_prt_bkg_value)/sum(aggr.prt$tot_prt_bkg_value)
aggr.prt$id <- seq(nrow(aggr.prt))
num.top.partners<- sum(aggr.prt$per_tot_prt_bkg_value<.80)
top.partners50 <- sum(aggr.prt$per_tot_prt_bkg_value<.50)
```

As shown in the cumulative distribution below, `r top.partners50` partners and  `r num.top.partners` partners concentrate 50\% and 80\% respectively of the value generated to Expedia.

> **Result:** `r top.partners50` partners and  `r num.top.partners` partners concentrate 50\% and 80\% respectively of the value generated to Expedia

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.prt, aes(x=id,y=per_tot_prt_bkg_value)) + geom_point() + xlab("# partners") + ylab("% total booking value")
```

#### Volume, value and competition

Figure below shows the relationship between the volume of the market (x-axis) and number of partners (y-axis)
for each market. The size of points is proportional to market size.
Figure shows that larger markets (mostly above 5000 bookings) attract a high number of partners.
As we will see in the next section, this does not mean that all these partners are actually generating revenue.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
# for each partner, check the actual number of
ggplot(aggr.mkt, aes(y=num_partners,x=num_bkgs)) + geom_point(aes(size=tot_mkt_bkg_value)) + ylab("# partners") + xlab("num bookings")
```

### PART 3. Market vs Partners:

#### In a given market, how many partners are generating a substantial booking value?

Figure below shows the proportion of *top partners* (partners contributing to to 80\% of the revenue in the market),
with respect to the the actual number of competing partners in a market (*num partners*) for the top 20 markets.
This figure clearly shows that the number of partners contributing to the market booking value is sensibly smaller than the overall number of partners competing in the market.
We can also see that 10 of these top 20 markets, have less 10 or less partners making a significant amount of money.
These means that certain partners are doing particularly well in these markets.

> **Result**: number of partners contributing to the market booking value is sensibly smaller than the overall number of partners competing in the market.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5.2, fig.height=4}
# select top 20 markets
top20.mkts <- arrange(aggr.mkt,desc(tot_mkt_bkg_value))$mkt[seq(20)]

top20.aggr.prt.mkt <- aggr.prt.mkt[aggr.prt.mkt$mkt %in% top20.mkts,]

top20stats <- ddply(top20.aggr.prt.mkt,.(mkt),function(df){
  df <- arrange(df ,desc(mkt_share))
  df$cum_mkt_share <- cumsum(df$mkt_share)
  top_partners <- length(which(df$cum_mkt_share<0.8))
  if(top_partners==0) top_partners <- 1
  num_partners <- nrow(df)
  total_mkt_bkg_value <- sum(df$tot_prt_mkt_bkg_value)
  data.frame(num_partners,top_partners,total_mkt_bkg_value)
})
ggplot(top20stats,aes(x=num_partners,y=top_partners))+geom_point(aes(size=total_mkt_bkg_value)) + xlab("# total partners") + ylab("# top partners")

```

#### How many markets are actually contributing to partner's revenue?

Similarly to the result above, figure below shows the proportion of *top markets* (market contributing to to 80\% of the revenue for a partner),
with respect to the the actual number of competing partners in a market (*num partners*) for the top 10 markets.
Again, the number of markets contributing to the partners booking value is sensibly smaller than the overall number of markets.

> **Result**: only a subset of the markets in which a partner is presentc create value for the partner.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5.2, fig.height=4}
# select top 20 markets
top10.prt <- arrange(aggr.prt ,desc(tot_prt_bkg_value))$partner_id[seq(10)]

top10.aggr.prt.mkt <- aggr.prt.mkt[aggr.prt.mkt$partner_id %in% top10.prt,]

top10stats <- ddply(top10.aggr.prt.mkt,.(partner_id),function(df){
  df <- arrange(df ,desc(prt_share))
  df$prt_share <- cumsum(df$prt_share)
  top_markets <- length(which(df$prt_share<0.8))
  if(top_markets==0) top_markets <- 1
  num_markets <- nrow(df)
  total_prt_bkg_value <- sum(df$tot_prt_mkt_bkg_value)
  data.frame(num_markets,top_markets,total_prt_bkg_value)
})
ggplot(top10stats,aes(x=num_markets,y=top_markets))+geom_point(aes(size=total_prt_bkg_value)) + xlab("# total markets") + ylab("# top markets")
```

### PART 3. Characterizing markets and partners


#### Booking value per partner and market

Figure below shows the average price distribution per partner and market. Distribution is bimodal
with a peak at 0 and another around 0.25.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.prt.mkt,aes(x=mean_bkg_value)) + geom_histogram()+ xlab("Average booking value per market")
```

Basic descriptive statistics below:

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
summary(aggr.prt.mkt$mean_bkg_value)
```

Markets and partners can be characterized based on their volume and booking value.
I first present an overview of markets and partners together, and then look at individual
markets by considering a subsample of the data that includes the Top
20 markets and the Top 10 partners alone.

#### Partners' and Markets' volume vs booking value

Figures below show the the volume and booking value for partners and markets.
Partners and markets that sell over the average booking of 0.26 are both interesting
opportunties for Expedia to grow its business.

**Recommendation**: focus on growing markets and partners with low volume and high booking value.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.prt,aes(x=mean_bkg_value,y=num_bkgs)) + geom_point() + xlab("Mean bkg value per partner") + ylab("Num bkgs per partner")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=4, fig.height=3}
ggplot(aggr.mkt,aes(x=mean_bkg_value,y=num_bkgs)) + geom_point() + xlab("Mean bkg value per market") + ylab("Num bkgs per market")
```

#### Top 10's partners volume and booking value for the Top 20 markets

Figure belows shows the volume and booking value of the top 10 partners
in each of the top 20 markets. It can be seen that both differences
in volume and booking value exist within each market.

As shown in the figure, we can identify opportunities for growth.
According to each market's booking value average,
we can divide partners between those with higher and lower booking value.
According to the overall volume, we can divide partners between
those that generate higher volume and low volume.

* partners in the low volume, higher value segment represent opportunities to grow in volume.

* partners in the high volume, higher value segment represent opportunities to grow in volume
(and maybe value).

* partners with low volume, low value can be helped to grow to higher values.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
top20.mkts <- arrange(aggr.mkt,desc(tot_mkt_bkg_value))$mkt[seq(20)]
top10.prts <- arrange(aggr.prt,desc(tot_prt_bkg_value))$partner_id[seq(10)]
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
sel.aggr.prt.mkt <- aggr.prt.mkt[aggr.prt.mkt$mkt %in% top20.mkts,]
sel.aggr.prt.mkt <- sel.aggr.prt.mkt[sel.aggr.prt.mkt$partner_id %in% top10.prts,]
ggplot(sel.aggr.prt.mkt,aes(x=mean_bkg_value,y=num_bkgs))+geom_point(aes(color=partner_id)) + xlab("mean_bkg_value") + ylab("num_bkgs")+facet_wrap(~mkt)
```

#### Top 20 markets volume and booking for the Top 10 partners

Figure belows shows the volume and booking value in each of the 20 markets
for each of the top 10 partners.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}
ggplot(sel.aggr.prt.mkt,aes(x=mean_bkg_value,y=num_bkgs))+geom_point(aes(color=mkt)) + xlab("mean_bkg_value") + ylab("num_bkgs")+facet_wrap(~partner_id)
```
